The illusory promise of the Aligned Rank Transform

### 2025-06-30 text/10-intro.qmd
many changes
preparing new version

However, given the significant influence of such milestone methods on our community's research outcomes, they warrant a closer scrutiny. Our results suggest that ART's popularity step largely from an ignorance of the method's assumptions. Researchers often treat ART as a general-purpose nonparametric alternative when ANOVA's assumptions are violated, overlooking the fact that ART's alignment mechanism introduces stricter assumptions than ANOVA itself, including a linear relationship between the response variable and the effects, equal variances, and continuous responses.

>>>>

However, given their substantial impact on research outcomes in our community, these widely adopted methods deserve closer scrutiny. As the first author previously argued [@tsandilas:2018], the HCI community often lacks the necessary expertise to critically assess the validity of statistical techniques. Our findings suggest that the widespread use of ART largely stems from a lack of awareness regarding its underlying assumptions. Many researchers treat ART as a general-purpose nonparametric alternative when ANOVA's assumptions are violated, overlooking the fact that ART’s alignment mechanism imposes even stricter assumptions -- including a linear relationship between the response and predictors, equal variances, and continuous data.

### 2025-06-30 text/20-background.qmd
preparing new version

![Citations of the CHI 2011 paper on ARTool [@wobbrock:2011] as shown in Google Scholar on 28 April 2024. We analyzed the 439 citing bibliographic entries of 2023 in more detail.](images/ARTool-citations-28-april-2024.png){width=80% #fig-ARTool-citations}

>>>>

![Citations of the CHI 2011 paper on ARTool [@wobbrock:2011] as shown in Google Scholar on 28 April 2024. We analyzed the 439 citing bibliographic entries of 2023 in more detail. *Note: Citations have continued to increase throughout the 2024–2025 period.*](images/ARTool-citations-28-april-2024.png){width=80% #fig-ARTool-citations}

### 2025-06-30 text/80-Recommandations.qmd
preparing new version

Our experimental results on individual Likert items with 5, 7, and 11 levels show that parametric ANOVA is robust for the analysis of main effects, as long as variances are equal. On the other hand, we observed that parametric ANOVA method often fails to correctly infer interactions when main effects become large, and the problem becomes more apparent when ordinal thresholds are not equidistant. INT appears to reduce the impact of this problem but without offering a satisfying solution. We also found that at least for 5-level items with flexible thresholds, all methods may lead to unprecise effect size estimates. In conclusion, while ANOVA seems to be a valid method for hypothesis testing even based on the analysis of individual Likert-item responses, we identified situations that require special attention. Unfortunately, nonparametric approaches may not necessarily work well in these situations.

In our prior research [@fages:2022], we applied ordered probit models [@Burkner:2019;@Christensen2023ordinal] for analyzing ordinal responses in two experimental studies. Since the HCI community is not familiar with these models, we found it necessary to provide motivation for their use and offer additional explanations to reviewers and readers on how to interpret results. These models offer numerous advantages, including their ability to support interval estimation and prediction in addition to null hypothesis testing. However, researchers are required to acquaint themselves with concepts of generalized mixed-effects models or Bayesian inference in order to use them correctly. We hope that our examples in the paper and the supplementary materials can inspire the community to give these methods more thoughtful consideration.

>>>>

Our experimental results on individual Likert items with 5, 7, and 11 levels show that parametric ANOVA is robust for the analysis of main effects, as long as variances are equal. On the other hand, we observed that parametric ANOVA method often fails to correctly infer interactions when main effects become large, and the problem becomes more apparent when ordinal thresholds are not equidistant. INT appears to reduce the impact of this problem but without offering a satisfying solution. We also found that at least for 5-level items with flexible thresholds, all methods may lead to imprecise effect size estimates. In conclusion, while ANOVA seems to be a valid method for hypothesis testing even based on the analysis of individual Likert-item responses, we identified situations that require special attention. Unfortunately, nonparametric approaches may not necessarily work well in these situations.

In our prior research [@fages:2022], we applied ordered probit models [@Burkner:2019;@Christensen2023ordinal] for analyzing ordinal responses in two experimental studies. Since the HCI community is not familiar with these models, we found it necessary to provide motivation for their use and offer additional explanations to reviewers and readers on how to interpret results. These models offer numerous advantages, including support for interval estimation and prediction in addition to null hypothesis testing. However, researchers are required to acquaint themselves with concepts of generalized mixed-effects models or Bayesian inference in order to use them correctly. They also come with certain limitations -- for example, the need to carefully specify priors to ensure convergence in certain scenarios, or their reliance on strict assumptions about populations variances and thresholds.

We hope that our examples in the paper and the supplementary materials can inspire the community to give these methods more thoughtful consideration.

### 2025-06-30 text/abstract.qmd
preparing new version

###### Results
Our results indicate that ART systematically confounds effects across all the experimental designs we examined. Specifically, an effect on one factor may cause ART to incorrectly detect effects on other factors or their interactions. We show that the method fails under various conditions, including skewed distributions, binomial distributions, Likert-type data with five or seven ordinal levels, distributions with unequal variances, and experimental designs with missing data.

Surprisingly, ART operates as expected only under normal distributions with equal variances. In all other scenarios, it performs less effectively than simpler rank transformations and, in many situations, such as with ordinal data, it is significantly worse than parametric ANOVA. Our analysis of real-world data further demonstrates that using ART often leads to incorrect conclusions. Conversely, the inverse normal transformation emerges as the most versatile and robust nonparametric method, although we identify situations in which this method can also inflate errors or distort effects.

In addition, our results show that testing interactions in the presence of multiple main effects is challenging for all methods. We raise awareness about previous warnings regarding the interpretation of *removable interactions*.

>>>>

###### Results
Our results indicate that ART fails under a range of discrete distributions, including binomial distributions, and Likert-type data with 5, 7, and 11 ordinal levels. In addition, ART systematically confounds effects in skewed distributions or distributions with unequal variances. We report results for various experimental designs, including between-subjects, within-subjects, and mixed designs with two or three factors.

Surprisingly, ART functions as intended only under the strict assumption of a linear model with continuous responses, equal variances, and no missing data. In all other scenarios, it performs worse than simpler rank transformation methods and, in many cases -- such as with ordinal data -- significantly worse than parametric ANOVA. Our analysis of real-world data further demonstrates that using ART often leads to incorrect conclusions. By contrast, the inverse normal transformation proves to be the most versatile and robust nonparametric method, though we also identify scenarios where it inflates errors or distort effects.

In addition, our results show that testing interactions in the presence of multiple main effects remains a challenge for all methods. We raise awareness about previous warnings regarding the interpretation of *removable interactions*.
